Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-17 paper2.qmd
section:issues; process:write,figures

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Probability
  - Directed Acyclic Graphs
  - Bayesian methods
  - Thurstonian model
  - Comparative judgement
  - Structural Causal Models
  - Statistical modeling
abstract: |
  (to do)
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

<!-- abstract -->
<!-- recording: Sven and Tine 24.11.25; time: 00:58:20 - 00:59:20 -->

@@

![Discriminal processes for two written texts](/images/png/discriminal_process.png){#fig-discriminal_process width=70%}

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

![Discriminal difference for two written texts](/images/png/discriminal_difference.png){#fig-discriminal_difference width=70%}

@@

<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:00:25 - 00:04:30 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:14:50 - 00:19:04 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:06:55 - 00:24:50 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->
<!-- recording: Sven and Tine 24.12.09; time: 00:03:20 - 00:08:10 -->

<!-- 1. What is the problem with the general form -->
Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form. Each case progressively incorporated additional simplifying assumptions into the model.

@@

<!-- Probability density and cumulative probability of the logistic and Normal distributions. Extracted from @Bramley_2008 [pp. 254-255]. -->
<!-- ::: -->

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic when researchers asesss complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- recording: Sven and Tine 24.12.09; time: 00:08:10 - 00:37:00 -->

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, the discriminal dispersions of stimuli play a critical role in determining the outcome of pairwise comparisons. Specifically, discrepancies in these dispersions shape the distribution of the discriminal difference, directly influencing the comparison outcome. @fig-dispersion illustrates this idea, assuming a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersion for Text A remains constant and that the texts have no correlation $(\rho=0)$.

@fig-dispersion reveals that more uncertainty in the trait perception of Text B compared to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.

![The discrepancy in the dispersions of stimuli and their effect on the distribution of the discriminal difference](/images/png/dispersion.png){#fig-dispersion width=70%}

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their discriminal processes [@Thurstone_1927a, pp. 373]. For example, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-dispersion when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The estimation accuracy of this discriminal difference, in turn, ensures reliable discriminal process estimates for the texts (citation needed?). It then intuitively follows that the outcome's ability to represent the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Then, incorrectly assuming equal dispersions between stimuli can lead the BTL model to introduce various statistical and measurement issues. For example, the model could overestimate the accuracy of the outcome in reflecting the "true" discriminal differences between stimuli. This overestimation may result in spurious inferences about these differences [@McElreath_2020, pp. 370] and, by extension, about the stimuli's discriminal processes. @fig-dispersion also illustrates this scenario when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows any discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the common practice in CJ literature of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] may inadvertently discard valuable information, introducing bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. These biases are often unpredictable, as they depend on the specific stimuli excluded from the analysis.

@@

@@

<!-- 1. What does the zero correlation between stimuli imply? -->
Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another. Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons. @fig-correlation illustrates this concept, assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersions for both texts remain constant.

@@

![The correlation between stimuli and its effect on the distribution of the discriminal difference](/images/png/correlation.png){#fig-correlation width=70%}

<!-- 2. But this is not the scenario we observe -->
However, in experimental practice, researchers typically follow this process in reverse. For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts (citation needed?).

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->

@@

<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need? Who should judge? How many judgments should they make? And crucially, how can we be confident that they are basing their judgments on acceptable criteria? If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld. We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used. However, work to derive some general principles for good practice would be valuable.[@Kelly_et_al_2022, pp. 683] -->

>>>>

---
title: "Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation"
author:
  - name: 
      given: Jose Manuel
      family: Rivera Espejo
    orcid: 0000-0002-3088-2783
    url: https://www.uantwerpen.be/en/staff/jose-manuel-rivera-espejo_23166/
    email: JoseManuel.RiveraEspejo@uantwerpen.be
    corresponding: true
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Tine
      family: van Daal
      non-dropping-particle: van
    orcid: https://orcid.org/0000-0001-9398-9775
    url: https://www.uantwerpen.be/en/staff/tine-vandaal/
    email: tine.vandaal@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Sven
      family: De Maeyer
      non-dropping-particle: De
    orcid: 0000-0003-2888-1631
    url: https://www.uantwerpen.be/en/staff/sven-demaeyer/
    email: sven.demaeyer@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Training and education sciences
        group: Edubron
  - name: 
      given: Steven
      family: Gillis
    orcid: 
    url: https://www.uantwerpen.be/nl/personeel/steven-gillis/
    email: steven.gillis@uantwerpen.be
    corresponding: false
    affiliation:
      - name: University of Antwerp
        department: Linguistics
        group: Centre for computational linguistics, psycholinguistics, and sociolinguistics (CLiPS)
funding: 
  statement: "The project was founded through the Research Fund of the University of Antwerp (BOF)."
keywords:
  - Causal analysis
  - Directed Acyclic Graphs
  - Bayesian statistical methods
  - Thurstonian model
  - Comparative judgement
  - Probability
  - Statistical modeling
abstract: |
  This study revisits Thurstone's law of comparative judgments (CJ) by addressing two key limitations in traditional approaches. Firstly, it addresses the overreliance on the assumptions of Thurstone's Case V in the statistical analysis of CJ data. Secondly, it addresses the apparent disconnect between CJ's approach to trait measurement and hypothesis testing. We put forward a systematic approach based on causal analysis and Bayesian statistical methods, which results in a model that facilitates a more comprehensive understanding of the factors influencing CJ experiments while offering a robust statistical translation. The new model accommodates unequal dispersions and correlations between stimuli, enhancing the reliability and validity of CJ's trait estimation, thereby ensuring the accurate measurement and interpretation of comparative data. The paper highlights the relevance of this updated framework for modern empirical research, particularly in education and social sciences. This contribution advances current research methodologies, providing a robust foundation for future applications in diverse fields.
key-points:
  - (to do)
date: last-modified
bibliography: references.bib
notebook-links: global
lightbox: true
citation: true
  # type: article-journal
  # container-title: "Psychometrika"
  # doi: ""
  # # url: https://example.com/summarizing-output
execute: 
  cache: true
  # eval: true
  echo: false
  # output: true
  # include: true
  warning: false
  error: false
  message: false
---

<!-- abstract -->
<!-- recording: Sven and Tine 24.11.25; time: 00:58:20 - 00:59:20 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:11:50 - 00:26:10, (Here we did it!) -->

@@

::: {#fig-thurstone_theory layout-ncol=2}

![Discriminal processes](/images/png/discriminal_process.png){#fig-discriminal_process width=100%}

![Discriminal difference](/images/png/discriminal_difference.png){#fig-discriminal_difference width=100%}

Hypothetical discriminal processes and discriminant difference along a quality trait continuum for two written texts.
:::

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference illustrates the distribution of the discriminal difference for the hypothetical texts depicted in panel @fig-discriminal_process. The figure indicates that the judge perceives Text B as having significantly higher quality than Text A. This conclusion rests on two key observations: the positive difference between their modal discriminal processes $(S_{B} - S_{A} > 0)$ and the probability area where the discriminal difference distinctly favors Text B over Text A, represented by the shaded gray area denoted as $P(B > A)$. As a result, the dichotomous outcome of this comparison is more likely to favor Text B over Text A.

@@

<!-- recording: Sven and Tine 24.10.25; time: 00:31:10 - 00:52:20 -->
<!-- recording: Sven and Tine 24.11.12; time: 00:00:00 - 00:00:00 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:00:25 - 00:04:30 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:14:50 - 00:19:04 -->
<!-- recording: Sven and Tine 24.11.18; time: 00:21:37 - 00:47:20 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:06:55 - 00:24:50 -->
<!-- recording: Sven and Tine 24.11.25; time: 00:35:10 - 00:49:30 -->
<!-- recording: Sven and Tine 24.12.09; time: 00:03:20 - 00:08:10 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:00:00 - 00:06:10 -->

<!-- 1. What is the problem with the general form -->
Thurstone observed that the general form of the theory, outlined in @sec-thurstone_theory, created a trait scaling problem. The model required estimating more "unknown" parameters than the available pairwise comparisons [@Thurstone_1927b, pp. 267]. To address this issue and facilitate the practical application of the theory, he developed five cases derived from this general form, each case progressively incorporating additional simplifying assumptions into the model.

@@

<!-- Probability density and cumulative probability of the logistic and Normal distributions. -->
<!-- ::: -->

<!-- 4. But what are limitations of Case V (and the BTL model)? -->
However, Thurstone originally developed Case V to provide a "rather coarse scaling" of traits [@Thurstone_1927b, pp. 269], prioritizing statistical simplicity over precision in trait measurement [@Kelly_et_al_2022, pp. 677]. He explicitly warned against its untested application, stating that its use "should not be made without (an) experimental test" [@Thurstone_1927b, pp. 270], acknowledging that some assumptions could prove problematic when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376]. Consequently, given that modern CJ applications frequently involve such traits and stimuli, two main assumptions of Case V and, by extension, of the BTL model may not consistently hold in theory or practice: the assumption of equal dispersion and zero correlation between stimuli.

@@

<!-- recording: Sven and Tine 24.12.09; time: 00:08:10 - 00:37:00 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:06:10 - 00:08:50 -->

<!-- 1. What does the equal dispersion between stimuli imply? -->
According to the theory, discrepancies in the discriminal dispersions of stimuli shape the distribution of the discriminal difference, exerting a direct influence on the outcome of pairwise comparisons. @fig-dispersion illustrates this idea using a thought experiment in which a researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. Furthermore, the figure assumes that the discriminal dispersion for Text A remains constant and that the texts are uncorrelated $(\rho=0)$. The figure reveals that an increase in the uncertainty associated with the perception of Text B in comparison to Text A, $(\sigma_{B}-\sigma_{A})$, broadens the distribution of their discriminal difference. This broadening affects the probability area where the discriminal difference distinctly favors Text B over Text A, expressed as $P(B > A)$, ultimately influencing the comparison outcome. Additionally, the figure reveals that when the discriminal dispersions of the texts are equal $(\sigma_{B}-\sigma_{A}=0)$, the discriminal difference is more likely to favor Text B over Text A (shaded gray area), compared to situations where their dispersions differ.

<!-- 2. But this is not the scenario we observe -->
In experimental practice, however, this process occurs in reverse. Researchers first observe the comparison outcome and then use the BTL model to infer the discriminal difference between the stimuli and their respective discriminal processes [@Thurstone_1927a, pp. 373]. For instance, when researchers observe a large sample of outcomes favoring Text B over Text A and correctly assume equal dispersions between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated in @fig-dispersion, where the model's discriminal difference distribution aligns with the "true" distribution, represented by the thick continuous line corresponding to $\sigma_{B}-\sigma_{A}=0$. The accuracy of this estimated discriminal difference ensures reliable estimates of the texts’ discriminal processes [(citation needed?){style="color:red;"}. Therefore, it is reasonable to assume that the outcome's ability to reflect the "true" differences between stimuli largely depends on the validity of the model's assumptions, particularly the assumption of equal dispersions.

@@

<!-- 4. So, what happens when the assumption does not hold? -->
Therefore, incorrectly assuming equal dispersions between stimuli is not without harm, as it can cause the BTL model to introduce significant statistical and measurement issues. For instance, the model may overestimate how accurately the outcome reflects the "true" discriminal differences between stimuli. This overestimation can lead researchers to draw spurious conclusions about these differences [@McElreath_2020, pp. 370] and, by extension, of the stimuli underlying discriminal processes. @fig-dispersion illustrates this issue when the model's discriminal difference distribution aligns with the thick continuous line for $\sigma_{B}-\sigma_{A}=0$, while the "true" discriminal difference follows a discontinuous line where $\sigma_{B}-\sigma_{A} \neq 0$. Moreover, if researchers recognize that misfit statistics highlight critical differences in dispersions, the usual CJ practice of excluding stimuli based on these statistics [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018] can unintentionally discard valuable information. This exclusion can introduce bias into trait estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are often unpredictable, as they depend on which stimuli are excluded from the analysis.

@@

::: {#fig-caseV_issues layout-ncol=2}

![Discrepancies in the dispersions of stimuli](/images/png/dispersion.png){#fig-dispersion width=100%}

![Correlation between stimuli](/images/png/correlation.png){#fig-correlation width=100%}

The effect of dispersion discrepancies and stimulus correlation on the distribution of the discriminal difference.
:::

@@

<!-- recording: Sven and Tine 24.12.17; time: 00:08:50 - 00:10:30 -->

<!-- 1. What does the zero correlation between stimuli imply? -->
Denoted by $\rho$, the correlation measures the dependence of a judge's perception of the trait in one stimulus on his perception of the same trait in another. Like the discriminal dispersions, this correlation shapes the distribution of the discriminal difference and directly influences the outcomes of pairwise comparisons. @fig-correlation illustrates this concept, again assuming the researcher can observe the discriminal processes for the texts shown in @fig-discriminal_process. The figure also considers that the discriminal dispersions for both texts remain constant.

@@

<!-- 2. But this is not the scenario we observe -->
However, in experimental practice, researchers typically follow this process in reverse. For example, when they observe a large sample of outcomes favoring Text B over Text A and correctly assume zero correlation between the texts, the BTL model estimates a discriminal difference distribution that accurately represents the "true" discriminal difference of the texts. This scenario is illustrated with @fig-correlation when the discriminal difference distribution of the model aligns with the "true" distribution, represented by the thick continuous line corresponding to $\rho=0$. The accuracy of this discriminal difference estimation, in turn, ensures reliable estimates for the discriminal process of the texts [(citation needed?){style="color:red;"}.

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->

@@

<!-- The field of comparative judgment, while growing, is still relatively small, and at present we do not have compelling answers to fundamental questions such as: how many judges do we need? Who should judge? How many judgments should they make? And crucially, how can we be confident that they are basing their judgments on acceptable criteria? If judges are not using construct-relevant criteria, then the argument that comparative judgments are necessarily valid cannot be upheld. We recognise that there is unlikely to be a single answer to these questions for all the cases in which comparative judgment may be used. However, work to derive some general principles for good practice would be valuable. [@Kelly_et_al_2022, pp. 683] -->
